$\pi$ with leftovers: \\ a mechanisation in Agda

### 2020-05-04 paper/paper.tex
Minor stuff

  \todo{we haven't introduced the forall and implicit syntax}
  \todo{introduce DEC and exists}

  \begin{figure}[h]
  \begin{equation*}
  \begin{aligned}
    &\zero                  &:{} &                      &        & C \\
    &\one                   &:{} &                      &        & C \\
    &\op{\_}{\_}{\_}        &:{} &                      &        & C \to C \to C \to \Set \\
    &\func{\cdot-compute}  &:{} &\forall y z           & \to \; & \type{DEC} \; (\type{\exists} x . \; (\op{x}{y}{z})) \\
    &\func{\cdot-unique}   &:{} &\forall \{x x' y z\}  & \to \; & \op{x'}{y}{z} \to \op{x}{y}{z} \to x' \equiv x \\
    &\func{\cdot-unique^l} &:{} &\forall \{x y y' z\}  & \to \; & \op{x}{y'}{z} \to \op{x}{y}{z} \to y' \equiv y \\
    &\func{0\cdot-min^l}   &:{} &\forall \{y z\}       & \to \; & \op{\zero}{y}{z} \to y \equiv \zero \\
    &\func{\cdot-id^l}     &:{} &\forall \{x\}         & \to \; & \op{x}{\zero}{x} \\
    &\func{\cdot-comm}     &:{} &\forall \{x y z\}     & \to \; & \op{x}{y}{z} \to \op{x}{z}{y} \\
    &\func{\cdot-assoc}    &:{} &\forall \{x y z u v\} & \to \; & \op{x}{y}{z} \to \op{y}{u}{v} \to \type{\exists} w . \; (\op{x}{u}{w} \times \op{w}{v}{z})
  \end{aligned}
  \end{equation*}
  \caption{Algebra $\Algebra_C$ algebra on a partial commutative monoid.}
  \label{fig:multiplicities}
  \end{figure}
\end{nidefinition}

@@

\begin{nidefinition}
  We use indexed sets of usage algebras to allow several usage algebras to coexist in our type system.
  An indexed set of usage algebras is a nonempty type $\Idx$ of indices together with an interpretation $\Usage_{idx}$ of indices into types, and an interpretation of indices into usage algebras.
  $\Idx$ must be non-empty.
  Refer to \autoref{fig:indexed-multiplicities} for a formal definition.
  \begin{figure}[h]
    \begin{equation}
      \begin{split}
        &\Idx               &:{} &\Set \\
        &\type{\exists IDX} &:{} &\Idx \\
        &\Usage             &:{} &\Idx \to \Set \\
        &\type{ALGEBRAS}    &:{} &(idx : \Idx) \to \Algebra_{\Usage_{idx}}
      \end{split}
    \end{equation}
    \caption{Indexed set of usage algebras.}
    \label{fig:indexed-multiplicities}
  \end{figure}

@@

\begin{nidefinition}
  \autoref{fig:prectx} defines types as either unit types (rule $\unit$), base types (rule $\base{n}$) or channel types (rule $\channel{t}{x}$).
  The unit type $\unit$ serves as a proof of inhabitance for types.
  The base type $\base{n}$ uses natural numbers as placeholders for types, the host language can then interpret the former into the latter --- e.g. $0$ as booleans, $1$ as natural numbers.
  This is for convenience, so that there is less universe polymorphism polluting specifications.
  The type $\channel{t}{x}$ of a channel determines what type of data ${t}$ and what usage annotations ${x}$ are sent over that channel.
  This notation is similar to Kobayashi's $[t]\mathbf{chan}_{(i^{y},o^{z})}$, where $y,z$ are the multiplicities of input and output \cite{K07}.
  We keep track of types in a length-indexed $\PreCtx_n$, defined by the last row of rules in \autoref{fig:prectx}.

@@

\todo{Maybe fig 6 can go in appendix in order to make room for 12 and 13 to come back? This whole paragraph from def 6 below could be cut and brought to app? and only commented upon}

@@

\begin{example}
  \label{example-derivations}
  Let us type a derivation of the process $\func{p}$ using linear and shared types. 
  Let $\omega$ be a shortcut for $(\one \comma \one)$ in an algebra of shared types.
  Let $\epsilon$ be the empty usage context --- where all usage annotations are $(\zero \comma \zero)$.
  To type $\func{p}$, we need a channel over which we can transmit another channel.
  We therefore create a new channel with usage annotations $(\one \comma \one)$ capable of transiting data of type $\channel{\unit}{\omega}$, and declare (with $\li$) that we will only transmit its input capability.
  The left-hand process is typed following its syntax.
  On the right-hand, we need the channel that is going to be sent.
  The type $\unit$ of this channel is entirely determined by the type of the channel over which it is going to be transmited.
  Its usage annotation is $(\one \comma \one)$: we will transmit $\li$ and save $\lo$ for the continuation.

@@

  $\func{\_}$ : $\types{\constr{[]} \comma \unit}{\constr{[]} \comma \omega}{\func{p}}{\epsilon}$
  $\func{\_}$ = $\constr{chan} \; \channel{\unit}{\omega} \; \li \; \one \; (\constr{comp}$
       $(\constr{recv} \; \constr{0} \; (\constr{recv} \; \constr{0} \; \constr{end}))$
       $(\constr{chan} \; \unit \; \omega \; \one \; (\constr{send} \; (\suc \; \constr{0}) \; \constr{0} \; (\constr{send} \; \constr{0} \; (\suc \; \suc \; \constr{0}) \; \constr{end})))
       )$
  \end{lstlisting}
\end{example}

@@

\begin{nitheorem}
  \label{thm:weakening}
  Inserting a new variable into the context preserves the well-typedness of a process as long as the usage annotation of the inserted variable is preserved as a leftover.

@@

\begin{nitheorem}
  \label{thm:strengthening}
  Removing an unused variable preserves the well-typedness of a process.

>>>>

  \begin{figure}[h]
  \begin{equation*}
  \begin{aligned}
    &\zero                  &:{} &                      &        & C \\
    &\one                   &:{} &                      &        & C \\
    &\op{\_}{\_}{\_}        &:{} &                      &        & C \to C \to C \to \Set \\
    &\func{\cdot-compute}  &:{} &\forall y z           & \to \; & \type{DEC} \; (\type{\exists} x . \; (\op{x}{y}{z})) \\
    &\func{\cdot-unique}   &:{} &\forall \{x x' y z\}  & \to \; & \op{x'}{y}{z} \to \op{x}{y}{z} \to x' \equiv x \\
    &\func{\cdot-unique^l} &:{} &\forall \{x y y' z\}  & \to \; & \op{x}{y'}{z} \to \op{x}{y}{z} \to y' \equiv y \\
    &\func{0\cdot-min^l}   &:{} &\forall \{y z\}       & \to \; & \op{\zero}{y}{z} \to y \equiv \zero \\
    &\func{\cdot-id^l}     &:{} &\forall \{x\}         & \to \; & \op{x}{\zero}{x} \\
    &\func{\cdot-comm}     &:{} &\forall \{x y z\}     & \to \; & \op{x}{y}{z} \to \op{x}{z}{y} \\
    &\func{\cdot-assoc}    &:{} &\forall \{x y z u v\} & \to \; & \op{x}{y}{z} \to \op{y}{u}{v} \to \type{\exists} w . \; (\op{x}{u}{w} \times \op{w}{v}{z})
  \end{aligned}
  \end{equation*}
  \caption{Algebra $\Algebra_C$ algebra on a partial commutative monoid.}
  \label{fig:multiplicities}
  \end{figure}
  \todo{we haven't introduced the forall and implicit syntax}
  \todo{introduce DEC and exists}
\end{nidefinition}

@@

\begin{nidefinition}
  We use indexed sets of usage algebras to allow several usage algebras to coexist in our type system.
  An indexed set of usage algebras is a nonempty type $\Idx$ of indices together with an interpretation $\Usage$ of indices into types, and an interpretation $\type{ALGEBRAS}$ of indices into usage algebras (\autoref{fig:indexed-multiplicities}).
  \begin{figure}[h]
    \begin{equation}
      \begin{split}
        &\Idx               &:{} &\Set \\
        &\type{\exists IDX} &:{} &\Idx \\
        &\Usage             &:{} &\Idx \to \Set \\
        &\type{ALGEBRAS}    &:{} &(idx : \Idx) \to \Algebra_{\Usage_{idx}}
      \end{split}
    \end{equation}
    \caption{Indexed set of usage algebras.}
    \label{fig:indexed-multiplicities}
  \end{figure}

@@

\begin{nidefinition}
  \autoref{fig:prectx} defines types as either unit types (rule $\unit$), base types (rule $\base{n}$) or channel types (rule $\channel{t}{x}$).
  The unit type $\unit$ serves as a proof of inhabitance for types.
  The base type $\base{n}$ uses natural numbers as placeholders for types (the host language can then interpret the former into the latter --- e.g. $0$ as booleans, $1$ as natural numbers) so that we avoid having to deal with universe polymorphism.
  The type $\channel{t}{x}$ of a channel determines what type of data ${t}$ and what usage annotations ${x}$ are sent over that channel.
  This notation is similar to Kobayashi's $[t]\mathbf{chan}_{(i^{y},o^{z})}$, where $y,z$ are the multiplicities of input and output \cite{K07}.
  We keep track of types in a length-indexed $\PreCtx_n$, defined by the last row of rules in \autoref{fig:prectx}.

@@

@@

\begin{example}
  \label{example-derivations}
  Let us type a derivation of the process $\func{p}$ using linear and shared types. 
  Let $\func{\omega}$ be a shortcut for $(\one \comma \one)$ in an algebra of shared types.
  Let $\epsilon$ be the empty usage context --- where all usage annotations are $(\zero \comma \zero)$.
  To type $\func{p}$, we need a channel over which we can transmit another channel.
  We therefore create a new channel with usage annotations $(\one \comma \one)$ capable of transiting data of type $\channel{\unit}{\func{\omega}}$, and declare (with $\li$) that we will only transmit its input capability.
  The left-hand process is typed following its syntax.
  On the right-hand, we need the channel that is going to be sent.
  The type $\unit$ of this channel is entirely determined by the type of the channel over which it is going to be transmited.
  Its usage annotation is $(\one \comma \one)$: we will transmit $\li$ and save $\lo$ for the continuation.

@@

  $\func{\_}$ : $\types{\constr{[]} \comma \unit}{\constr{[]} \comma \func{\omega}}{\func{p}}{\epsilon}$
  $\func{\_}$ = $\constr{chan} \; \channel{\unit}{\func{\omega}} \; \li \; \one \; (\constr{comp}$
       $(\constr{recv} \; \constr{0} \; (\constr{recv} \; \constr{0} \; \constr{end}))$
       $(\constr{chan} \; \unit \; \func{\omega} \; \one \; (\constr{send} \; (\suc \; \constr{0}) \; \constr{0} \; (\constr{send} \; \constr{0} \; (\suc \; \suc \; \constr{0}) \; \constr{end})))
       )$
  \end{lstlisting}
\end{example}

@@

\begin{nitheorem}
  \label{thm:weakening}
  Weakening captures the idea that inserting a new variable into the context preserves the well-typedness of a process as long as the usage annotation of the inserted variable is preserved as a leftover.

@@

\begin{nitheorem}
  \label{thm:strengthening}
  Strengthening captures the idea that removing an unused variable preserves the well-typedness of a process.
